---
title: "BACE: Bayesian Phylogenetic and Correlated Trait Imputation"
date: "`r Sys.Date()`"
author: "Shinichi Nakagawa, Szymek Drobniak, Daniel Noble"
bibliography: ../bib/BACE.bib
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    toc-title: "**Table of Contents**"
    output-file: "index.html"
    theme: cosmo
    embed-resources: true
    code-fold: show
    code-tools: true
    number-sections: true
    fontsize: "12"
    max-width: "10"
    code-overflow: wrap
crossref: 
  fig-title: Figure     # (default is "Figure")
  tbl-title: Table     # (default is "Table")
  title-delim: —     # (default is ":")
  fig-prefix: Fig.   # (default is "Figure")
  tbl-prefix: Tab.    # (default is "Table")
editor_options: 
  chunk_output_type: console
---

# Introduction
We provide a brief introduction to the BACE package for Bayesian phylogenetic and correlated trait imputation. The BACE package allows users to impute missing data in comparative datasets using Bayesian methods that account for phylogenetic relationships among species.

# Installation
You can install the BACE package from GitHub using the following command:

```{r}
#| label: install_bace
#| ehcho: true

# Install from github and load
#install.packages("pacman")
  pacman::p_load(devtools)
devtools::install_github("daniel1noble/BACE")
  pacman::p_load(BACE, ape, phytools, MCMCglmm, dplyr, magrittr)
```

If you encounter any issues during installation, please ensure that you have the required dependencies installed.

# Documentation
The main function in the BACE package is `bace_imp()`, which performs Bayesian imputation using chained equations. You can access the documentation for this function using the following command:
```{r}
#| label: bace_imp_doc
#| echo: true
?bace_imp
```

As you will see, the function allows you to specify fixed effects, random effects, phylogenetic trees, and other parameters for the imputation process. At the bare minimum, you need to provide a formula for the fixed effects, a formula for the random effects (including phylogeny), a phylogenetic tree, and the dataset with missing values.

# Example Usage

We'll first simulate a dataset with missing values and a phylogenetic tree. We can do this using the `sim_bace()` function.

```{r}
#| label: simulate_data
#| echo: true

set.seed(123)
# Simulate data. Note that threshold4 indicates an ordinal variable with 4 levels, and multinomial3 indicates a categorical variable with 3 levels. You can replace the numbers 
sim_data <- sim_bace(response_type = "poisson",
                   predictor_types = c("gaussian", "binary", "threshold4", "multinomial3"),
                      phylo_signal = c(0.55, 0.75, 0.8, 0.7, 0.7),
                       missingness = c(0.2, 0.3, 0.4, 0.2, 0.5),
                                rr = TRUE,
                           rr_form = list(species = c("x1")))
```

We can then view the simulated data and phylogenetic tree because the `sim_bace()` function returns a list containing both objects. It also has a lot of other useful information about the simulation users can explore.

```{r}
#| label: view_simulated_data
#| echo: true

# Object contains both the simulated data and phylogeny
    data <- sim_data$data
    tree <- sim_data$tree

# View first few rows of the data
head(data)
```


Here, we can see that the dataset contains lots of missing values across different variable types. Of course, with normal modelling, we would need to exclude these cases to use a complete datset. However, with BACE we simply feed in the entire dataset, specify the models we want to use for imputation (or the real model of interest), and impute the missing data. 

# Using BACE for Imputation
Now that we have the data we can now use the `bace_imp()` function to impute the missing data.

```{r}
#| label: bace_imputation
#| echo: true 

# First check that all variables are correctly classified. Correctly. If using sim_bace() this is done automatically, but always worth checking.
  str(data)

# Perform BACE imputation. NOTE: YOU MUST HAVE A MODEL FOR EACH MISSING VARIABLE IN THE DATA IF YOU PROVIDE A LIST. This is not the case if you provide a single formula because BACE will internally build models for the other variables based on their types.
bace_impute <- bace_imp(fixformula = list("y ~ x1 + x2",
                                          "x1 ~ x2",
                                          "x2 ~ x4",
                                          "x3 ~ x1 + x2",
                                          "x4 ~ x1 + x2"),
                    ran_phylo_form = "~ 1 |species",
                             phylo = tree,
                              data = data,
                              runs = 5)
                       
```

If you want to just model your main variables and ignore others you can also simply feed in a single formula as below. Here, BACE will automatically build models for the other variables based on their types and the variables present in the dataset.

```{r}
#| label: bace_imputation_simple
#| echo: true

# Fit BACE with a single formula
bace_impute2 <- bace_imp(fixformula = "y ~ x1 + x2",
                     ran_phylo_form = "~ 1 |species",
                              phylo = tree,
                               data = data,
                               runs = 5, nitt = 20000, burnin = 5000, thin = 10)     
```