---
title: "BACE simulation study: does it recover the truth under missing data?"
date: 2026-08-08
format:
  html:
    embed-resources: true
    html-math-method: katex
    toc: true
    toc-depth: 2
    number-sections: true
    theme: cosmo
    fig-width: 8
    fig-height: 4.4
    code-fold: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(ggplot2)
OUT <- "."   # rendered from dev/simulation_results/
A <- readRDS(file.path(OUT, "recovery_simulation.rds"))       # Study A (post-fix)
B <- readRDS(file.path(OUT, "response_type_simulation.rds"))  # Study B (post-fix)
Ap <- A$per_rep
Bp <- B$per_rep[B$per_rep$status == "ok", ]
A_REPS <- A$cfg$reps
B_REPS <- B$cfg$reps
M_FINAL <- B$cfg$n_final

# Pre-fix (2026-07) Study B per-rep results, for the before/after comparison
ARCH <- file.path(OUT, "archive_2026-07_prebugfix", "response_type_per_rep.csv")
Bold <- if (file.exists(ARCH)) {
  z <- utils::read.csv(ARCH); z[z$status == "ok", ]
} else NULL

OK <- c(MCAR = "#56B4E9", MAR = "#D55E00")

# Monte Carlo standard errors
se_mean <- function(v) { v <- v[is.finite(v)]; sd(v) / sqrt(length(v)) }
se_prop <- function(v) { v <- v[is.finite(v)]; p <- mean(v); sqrt(p * (1 - p) / length(v)) }
med_se  <- function(v) { v <- v[is.finite(v)]; 1.2533 * mad(v) / sqrt(length(v)) }

# nominal-coverage Monte Carlo band
cov_band <- function(n) 1.96 * sqrt(0.95 * 0.05 / n)

theme_dw <- theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank(),
        plot.title = element_text(size = 12),
        legend.position = "top", legend.title = element_blank())

summ <- function(df, group, col, stat = c("mean","prop","median")) {
  stat <- match.arg(stat)
  do.call(rbind, lapply(split(df, list(df[[group]], df$mechanism), drop = TRUE),
    function(g) {
      v <- g[[col]]
      est  <- switch(stat, mean = mean(v[is.finite(v)]),
                     prop = mean(v[is.finite(v)]), median = median(v[is.finite(v)]))
      mcse <- switch(stat, mean = se_mean(v), prop = se_prop(v), median = med_se(v))
      data.frame(group = g[[group]][1], mechanism = g$mechanism[1],
                 est = est, lo = est - 1.96 * mcse, hi = est + 1.96 * mcse)
    }))
}
dwplot <- function(d, xlab, order, title) {
  d$group <- factor(d$group, levels = order)
  ggplot(d, aes(est, group, xmin = lo, xmax = hi, colour = mechanism)) +
    geom_pointrange(position = position_dodge(width = 0.5), fatten = 3) +
    scale_colour_manual(values = OK) +
    labs(x = xlab, y = NULL, title = title) + theme_dw
}
```

# What changed since the 2026-07 report {.unnumbered}

The 2026-07 run of this study surfaced poor interval coverage (gaussian ~0.76,
poisson ~0.69), catastrophic poisson imputations, and near-chance categorical
recovery. A four-way root-cause investigation (2026-08, adversarially
verified; full write-up in `.agents/investigation-2026-08.md`) found **six
defects** and fixed them:

1. **Improper final-phase MI** (the dominant bug): `bace_final_imp` fit every
   final model on the convergence chain's point imputations *as if observed* —
   response NAs are now reinstated so MCMCglmm does data augmentation.
2. **Finite-draw metric ceiling**: the 95% PI formed from the empirical
   2.5–97.5% quantiles of `n_final` draws covers at most $0.95\,(m-1)/(m+1)$
   (~0.845 at $m=15$) *even for a perfect imputer*; this report now also
   reports a ceiling-free $t$-interval (`cover95_t`) and probability-based
   discrete prediction sets (`cover95_prob`).
3. **Poisson point estimate** was a mean of lognormals (one chain excursion
   ⇒ million-count imputations, frozen everywhere by defect 1) — now a
   median-based rate with a data-adaptive guard (3·max observed + 5).
4. **Priors**: gaussian R back to Hadfield's (V=1, ν=0.002) — the 2026-04
   ν=2 setting *inflated* residual variance ~3× and biased λ̂ down; poisson
   gets log-link-scaled priors (R: V=0.1, ν=1; PX working prior α.V=1).
5. **Degenerate categorical DGP**: all K−1 liabilities shared one linear
   predictor (Bayes balanced-accuracy ceiling ≈ 0.50 — the old 0.43–0.46 was
   near-optimal, not a failure). The DGP now draws distinct per-liability
   slopes and random effects, and the per-replicate **Bayes ceiling** is
   reported alongside BACE's accuracy. The categorical default also switched
   to the true multinomial model (`ovr_categorical = FALSE`).
6. **Type detection**: integer variables were assigned poisson vs gaussian by
   a *marginal* AIC race that misfired exactly under strong phylogenetic
   signal — now deterministic (non-negative integers ⇒ poisson).

Pre-fix outputs are archived in `archive_2026-07_prebugfix/` and compared
against below (@sec-beforeafter).

# Aims

Evaluate whether **BACE recovers the truth** when data are missing: are its
imputed values and the parameters estimated from them **unbiased**, are its
uncertainty intervals **calibrated**, and does this hold **across response
distributions** and **missingness mechanisms**?

# What simulations we did

Two studies, both on `sim_bace()` with production MCMC
(`nitt` 4–5k, `burnin` 0.8–1k, `thin` 4–5, `n_final` `r M_FINAL`–20).

| Study | Response | Mechanisms | Replicates | Purpose |
|---|---|---|---|---|
| **A** | gaussian, known slope | MCAR, MAR | **`r A_REPS`** each | parameter recovery: slope bias + CI coverage, vs an oracle and complete-case |
| **B** | gaussian, poisson, binary, ordinal, categorical | MCAR, MAR | **`r B_REPS`** each | value recovery, bias, and interval calibration for every response type |

Study B is `r length(unique(Bp$type))` types × 2 mechanisms × `r B_REPS` reps =
**`r format(5*2*B_REPS, big.mark=",")` datasets** (`r nrow(Bp)` completed).

# Measures (what they mean)

For each replicate we hide 30% of the response, impute with BACE, and compare
to the known truth; averages over replicates carry **Monte Carlo standard
errors (MCSE)**.

- **Bias** (standardised): $\big(\overline{\hat y} - \overline{y}\big)/\mathrm{SD}(y)$
  on hidden cells (**0 = unbiased**); categorical uses total-variation distance
  of class proportions.
- **Coverage** — two estimators per family:
  - `cover95` *(quantile; kept for comparability with 2026-07)*: truth inside
    the empirical 2.5–97.5% range of the `n_final` draws. **Ceiling
    ≈ `r round(0.95*(M_FINAL-1)/(M_FINAL+1), 3)` at `r M_FINAL` draws even for
    a perfect imputer** — read it against that ceiling, not against 0.95.
  - `cover95_t` (continuous): $\bar y_{draws} \pm t_{m-1,.975}\, s_{draws}\sqrt{1+1/m}$
    — no finite-draw ceiling; target 0.95. `cover95_prob` (discrete): smallest
    class set reaching 0.95 cumulative probability under the models' averaged
    class-probability matrices; target ≥ 0.95.
- **Recovery**: correlation (continuous; plus Spearman and log1p-scale for
  poisson, since Pearson on raw counts is tail-dominated) or balanced accuracy
  (discrete). Categorical is judged **against its per-replicate Bayes
  ceiling** (`bayes_bal_acc`, computed from the true simulated class
  probabilities).

# Results

## Study A — parameter recovery (gaussian slope, `r A_REPS` reps)

```{r figA, fig.height=3.6}
mlab <- c(oracle="oracle", complete_case="complete-case", bace="BACE")
Ap$grp <- mlab[Ap$method]
cov <- summ(transform(Ap, mechanism=Ap$mechanism), "grp", "covered", "prop")
mb  <- summ(Ap, "grp", "mean_bias", "mean")
ord <- c("oracle","complete-case","BACE")
band <- cov_band(A_REPS)
p1 <- dwplot(cov, "95% CI coverage of true slope", ord, "Slope coverage") +
  annotate("rect", xmin=0.95-band, xmax=0.95+band, ymin=-Inf, ymax=Inf,
           alpha=0.12, fill="grey30") +
  geom_vline(xintercept=0.95, linetype=2) + xlim(0.7, 1.02)
p2 <- dwplot(mb, "bias in estimated mean of y", ord, "Marginal-mean bias") +
  geom_vline(xintercept=0, linetype=2)
patchwork::wrap_plots(p1, p2, nrow=1) & theme_dw
```

*Grey band = correct-calibration range at `r A_REPS` reps. Under MAR,
complete-case analysis is a biased sample of the response; BACE removes that
bias and its 95% CI covers the true slope at the nominal rate.*

```{r tabA}
agg <- do.call(rbind, lapply(split(Ap, list(Ap$grp, Ap$mechanism)), function(g)
  data.frame(Method=g$grp[1], Mech=g$mechanism[1],
    `Slope bias`=round(mean(g$slope_est)-0.8,3),
    `Coverage`=round(mean(g$covered),3),
    `Mean bias`=round(mean(g$mean_bias),3), check.names=FALSE)))
knitr::kable(agg[order(agg$Mech, agg$Method),], row.names=FALSE)
```

## Study B — recovery, bias, calibration by response type (`r B_REPS` reps each)

```{r figB_rec, fig.height=3.8}
cont <- c("poisson","gaussian"); disc <- c("categorical","binary","ordinal")
rc <- summ(Bp[Bp$type %in% cont,], "type", "recovery", "mean")
rd <- summ(Bp[Bp$type %in% disc,], "type", "balanced_accuracy", "mean")
q1 <- dwplot(rc, "correlation (imputed vs truth)", cont, "Continuous / count: recovery") + xlim(0,1)
q2 <- dwplot(rd, "balanced accuracy", disc, "Discrete: recovery") +
  geom_vline(xintercept=1/3, linetype=3, colour="grey50") + xlim(0.25,1)
# Bayes ceiling for categorical, if recorded
bc <- Bp$bayes_bal_acc[Bp$type == "categorical"]
if (any(is.finite(bc))) {
  q2 <- q2 + annotate("point", x = mean(bc[is.finite(bc)]), y = "categorical",
                      shape = 8, size = 3, colour = "grey30")
}
patchwork::wrap_plots(q1, q2, nrow=1) & theme_dw
```

*Dotted line = chance (1/3). Star = the categorical **Bayes-classifier
ceiling** given the DGP — balanced accuracy cannot exceed it.*

```{r figB_bias, fig.height=4}
allt <- c("categorical","ordinal","binary","poisson","gaussian")
bi <- summ(Bp, "type", "bias", "median")           # robust: poisson tail
Bp$cover_best <- ifelse(is.finite(Bp$cover95_t), Bp$cover95_t, Bp$cover95_prob)
cv <- summ(Bp, "type", "cover_best", "mean")
band <- cov_band(B_REPS)
b1 <- dwplot(bi, "median standardised bias (categorical: TV)", allt, "Bias") +
  geom_vline(xintercept=0, linetype=2)
b2 <- dwplot(cv, "95% interval (t) / probability-set coverage", allt,
             "Interval calibration (ceiling-free estimators)") +
  annotate("rect", xmin=0.95-band, xmax=0.95+band, ymin=-Inf, ymax=Inf,
           alpha=0.12, fill="grey30") +
  geom_vline(xintercept=0.95, linetype=2) + xlim(0.5,1.02)
patchwork::wrap_plots(b1, b2, nrow=1) & theme_dw
```

*Calibration uses the ceiling-free estimators: `cover95_t` for
gaussian/poisson, `cover95_prob` for binary/ordinal/categorical.*

```{r tabB}
mn <- function(x) round(mean(x, na.rm=TRUE), 3)
agg <- do.call(rbind, lapply(split(Bp, list(Bp$type, Bp$mechanism)), function(g)
  data.frame(Type=g$type[1], Mech=g$mechanism[1], n=nrow(g),
    Recovery=mn(g$recovery),
    `Rec (Spearman)`=mn(g$recovery_spearman),
    `Bal.acc`=mn(g$balanced_accuracy),
    `Bayes ceiling`=mn(g$bayes_bal_acc),
    `Median bias`=round(median(g$bias[is.finite(g$bias)]),3),
    `Cover (q)`=mn(g$cover95),
    `Cover (t)`=mn(g$cover95_t),
    `Cover (prob)`=mn(g$cover95_prob),
    `Runtime (s)`=round(mean(g$runtime_sec,na.rm=TRUE),1), check.names=FALSE)))
ord <- c("gaussian","poisson","binary","ordinal","categorical")
agg <- agg[order(match(agg$Type,ord), agg$Mech),]
knitr::kable(agg, row.names=FALSE)
```

*`Cover (q)` is the 2026-07-comparable quantile estimator — its perfect-imputer
ceiling at `r M_FINAL` draws is ≈ `r round(0.95*(M_FINAL-1)/(M_FINAL+1), 2)`,
so values near that ceiling indicate correct calibration.*

## Before vs after the 2026-08 fixes {#sec-beforeafter}

```{r beforeafter}
if (!is.null(Bold)) {
  key <- function(d, col) {
    x <- lapply(split(d, list(d$type, d$mechanism), drop=TRUE), function(g)
      data.frame(type=g$type[1], mech=g$mechanism[1],
                 val=round(mean(g[[col]], na.rm=TRUE), 3)))
    do.call(rbind, x)
  }
  ba <- Reduce(function(a,b) merge(a,b,by=c("type","mech")), list(
    setNames(key(Bold,"recovery"), c("type","mech","rec_pre")),
    setNames(key(Bp,  "recovery"), c("type","mech","rec_post")),
    setNames(key(Bold,"cover95"),  c("type","mech","covq_pre")),
    setNames(key(Bp,  "cover95"),  c("type","mech","covq_post")),
    setNames(key(Bp,  "cover95_t"),   c("type","mech","covt_post")),
    setNames(key(Bp,  "cover95_prob"),c("type","mech","covp_post"))))
  # catastrophe rate for poisson
  pcat <- function(d) with(d[d$type=="poisson" & is.finite(d$bias),],
                           tapply(abs(bias) > 1, mechanism, mean))
  ba <- ba[order(match(ba$type, ord), ba$mech),]
  names(ba) <- c("Type","Mech","Recovery (pre)","Recovery (post)",
                 "Cover q (pre)","Cover q (post)","Cover t (post)","Cover prob (post)")
  knitr::kable(ba, row.names=FALSE)
  cat(sprintf("\nPoisson catastrophe rate P(|std bias| > 1): pre MCAR %.3f / MAR %.3f  ->  post MCAR %.3f / MAR %.3f\n",
              pcat(Bold)["MCAR"], pcat(Bold)["MAR"], pcat(Bp)["MCAR"], pcat(Bp)["MAR"]))
} else {
  cat("Archived pre-fix per-rep data not found; before/after table skipped.")
}
```

*Note the categorical rows are not a like-for-like comparison: the 2026-08 DGP
is non-degenerate (per-liability slopes; Bayes ceiling reported above), whereas
the 2026-07 DGP capped balanced accuracy at ~0.50 by construction.*

## Timing

```{r figT, fig.height=3.2}
tm <- do.call(rbind, lapply(split(Bp, Bp$type), function(g)
  data.frame(group=g$type[1], mechanism="MCAR",
             est=mean(g$runtime_sec), lo=NA, hi=NA)))
tm$group <- factor(tm$group, levels=c("gaussian","poisson","binary","ordinal","categorical"))
ggplot(tm, aes(est, group)) +
  geom_point(colour="#0072B2", size=3) +
  geom_text(aes(label=sprintf("%.1fs", est)), hjust=-0.3, size=3.4) +
  labs(x="mean BACE runtime (s per dataset)", y=NULL,
       title="Timing by response type") + xlim(0, max(tm$est)*1.2) + theme_dw
```

*Reproduce:* `Rscript dev/12_recovery_simulation.R`,
`dev/13_response_type_simulation.R`, then render this file with
`quarto render dev/simulation_results/SIMULATION_REPORT.qmd`.
